---
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries, echo = FALSE}
library(caret)
library(e1071)
library(corrplot)
library(RColorBrewer)
```

### Datasets
Two sets of PS and NPS descriptors data were imported: one comprising the descriptors for all solvents (used both in the training and validation phase diagrams) and one comprising the descriptors for a reduced set of solvents (used only in phase diagrams that are in the training subset). The required preprocessing steps were determined using the latter training solvents set and applied on the former training/validation solvents set.
``` {r datasets}
PS_descriptors <- read.csv("Data/PS descriptors.csv")
NPS_descriptors <-read.csv("Data/NPS descriptors.csv")

PS_descriptors_training <- read.csv("Data/PS descriptors (training).csv")
NPS_descriptors_training <-read.csv("Data/NPS descriptors (training).csv")
```

### Removing descriptors containing missing data
``` {r check NA, echo = FALSE, results='hide'}
# Check whether the same number of columns containg missing data in training and training/validation dataset descriptors
if (ncol(PS_descriptors[,colSums(is.na(PS_descriptors)) != 0]) == ncol(PS_descriptors_training[,colSums(is.na(PS_descriptors_training)) != 0])) {
  print("True")
} else {
  print("False")
}
if (ncol(NPS_descriptors[,colSums(is.na(NPS_descriptors)) != 0]) == ncol(NPS_descriptors_training[,colSums(is.na(NPS_descriptors_training)) != 0])) {
  print("True")
} else {
  print("False")
}

# If both true, remove all variables containing a NA value
```
```{r PS remove NA}
PS_descriptors <-PS_descriptors[,colSums(is.na(PS_descriptors)) == 0]
PS_descriptors_training <- PS_descriptors_training[ , colSums(is.na(PS_descriptors_training)) == 0]
```
```{r, echo = FALSE}
cat("Number of PS descriptors removed:", 2702-length(colnames(PS_descriptors)),"\n")
```
```{r NPS remove NA}
NPS_descriptors <- NPS_descriptors[ , colSums(is.na(NPS_descriptors)) == 0]
NPS_descriptors_training <- NPS_descriptors_training[ , colSums(is.na(NPS_descriptors_training)) == 0]
```
```{r, echo = FALSE}
cat("Number of NPS descriptors removed:", 2702-length(colnames(PS_descriptors)),"\n")
```

### Zero and near-zero variance descriptors removal
The number of descriptors was reduced by removing all descriptors that comprise less than 2 unique values (i.e., zero variance descriptors) or 2 unique values when 1 of them is only present once (i.e., nero-zero variance descriptors). 
```{r PS zv and nzv}
PS_descriptors_nzv<-nearZeroVar(PS_descriptors_training, freqCut = 7, uniqueCut = 23)
PS_descriptors_training<-PS_descriptors_training[-c(PS_descriptors_nzv)]
PS_descriptors<-PS_descriptors[-c(PS_descriptors_nzv)]
```
``` {r, echo = FALSE}
cat("Number of PS descriptors removed:", length(PS_descriptors_nzv),"\n")
```
```{r NPS zv and nzv}
NPS_descriptors_nzv<-nearZeroVar(NPS_descriptors_training, freqCut = 11, uniqueCut = 16)
NPS_descriptors_training<-NPS_descriptors_training[-c(NPS_descriptors_nzv)]
NPS_descriptors<-NPS_descriptors[-c(NPS_descriptors_nzv)]
```
``` {r, echo = FALSE}
cat("Number of NPS descriptors removed:", length(NPS_descriptors_nzv),"\n")
```

### Descriptors transformation, scaling and centering
First, the skewness of the descriptors was estimated as:
```{r initial skeweness}
skewness_PS<-apply(PS_descriptors_training[,2:641],2,skewness)
skewness_NPS<-apply(NPS_descriptors_training[,2:684],2,skewness)
```
The results were plotted as histograms to visually access the overall level of the skewness in the solvent descriptors datasets for polar solvents and non-polar solvents.

```{r initial PS skewness, echo = FALSE, fig.width = 12,fig.height = 4, fig.align="center"}
ggplot(as.data.frame(skewness_PS), 
       aes(skewness_PS)) + 
  geom_histogram(binwidth = 0.2) +
  labs(title = "Polar solvent descriptors skewness", x="Lambda", y="Number of parameters") + 
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r initial NPS skewness, echo = FALSE,fig.width = 12,fig.height = 4, fig.align="center"}
ggplot(as.data.frame(skewness_NPS), 
       aes(skewness_NPS)) + 
  geom_histogram(binwidth = 0.2) + 
  labs(title = "Non-polar solvent descriptors skewness",x="Lambda", y="Number of parameters") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```
The NPS descriptors are notably more skewed than the PS ones.

To reduce the descriptors skewness the Yeo-Johnson transformation (which performs better than the alternative Box-Cox transformation at normalising variables containing zeros and negative numbers, such as the ones here) was applied to all predictors. \
Additionally, all predictors were scaled and centered, which is important for predictive model development where the best model is determined by calculating the difference between classifiers and new samples and ranking it according to scale, such as support vector machines.
```{r preprocessing determination}
PS_preprocessing <- preProcess(PS_descriptors_training[,2:641], method = c("scale", "center", "YeoJohnson"))
NPS_preprocessing <- preProcess(NPS_descriptors_training[,2:684], method = c("scale", "center", "YeoJohnson"))
```

``` {r PS preprocessing, echo = FALSE}
cat("Polar solvent descriptors transormation \n")
PS_preprocessing
```

``` {r NPS preprocessing, echo = FALSE}
cat("Non-polar solvent descriptors transormation \n")
NPS_preprocessing
```

```{r preprocessing application, echo = FALSE}
PS_descriptors[,2:641] <- predict(PS_preprocessing, PS_descriptors[,2:641])
PS_descriptors_training[,2:641] <- predict(PS_preprocessing, PS_descriptors_training[,2:641])
NPS_descriptors[,2:684] <- predict(NPS_preprocessing, NPS_descriptors[,2:684])
NPS_descriptors_training[,2:684] <- predict(NPS_preprocessing, NPS_descriptors_training[,2:684])
```
After the transformation, the overall skewness of the descriptors was significantly reduced.

```{r final skewness, echo=FALSE}
skewness_PS_2<-apply(PS_descriptors_training[,2:641],2,skewness)
skewness_NPS_2<-apply(NPS_descriptors_training[,2:684],2,skewness)
```

```{r final PS skewness, echo = FALSE, fig.width = 12,fig.height = 4, fig.align="center"}
ggplot(as.data.frame(skewness_PS_2), aes(skewness_PS_2)) + 
  geom_histogram(binwidth = 0.2) + 
  labs(title = "Polar solvent descriptors skewness after transoformation", x="Lambda", y="Number of parameters") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r final NPS skewness, echo = FALSE, fig.width = 12,fig.height = 4, fig.align="center"}
ggplot(as.data.frame(skewness_NPS_2), aes(skewness_NPS_2)) + 
  geom_histogram(binwidth = 0.2) + 
  labs(title = "Non-polar solvent descriptors skewness after transoformation", x="Lambda", y="Number of parameters") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

### Cross-correlations removal

A number of predictive models, especially linear models (e.g., logistic regression and linear discriminant analysis), require that the predictors used are not co-linear to avoid errors in estimating the contribution (weight) of each predictor to the model.

Hence,the cross-correlation between the predictors was calculated to assess the presence of co-linearity in the predictor data set.

<center> 
**Correlation plot for PS descriptors**
</center>

```{r PS corrplot, echo=FALSE, fig.width = 10,fig.height = 10, fig.align="center"}
PS_corr <- cor(PS_descriptors_training[,2:641], method="pearson")
cols <- c(rev(brewer.pal(9, "Blues")),brewer.pal(9, "Reds"))
corrplot::corrplot(PS_corr,
                   order = "hclust",
                   tl.cex = 0.2,
                   addgrid.col = rgb(1,1,1,.01),
                   col = colorRampPalette(cols)(51))
```
<center>
**Correlation plot for NPS descriptors**
</center>
```{r NPS corrplot, echo=FALSE, fig.width = 10,fig.height = 10, fig.align="center"}
NPS_corr <- cor(NPS_descriptors_training[,2:684], method="pearson")
cols <- c(rev(brewer.pal(9, "Blues")),brewer.pal(9, "Reds"))
corrplot::corrplot(NPS_corr,
                   tl.cex = 0.2,
                   order = "hclust",
                   addgrid.col = rgb(1,1,1,.01),
                   col = colorRampPalette(cols)(51))
```

As can be seen from the two correlation plots above, there is significant co-linearity in both descriptor datasets.

To reduce this co-linearity a Pearson two-tailed correlation test was used. A Pearson correlation of 0.735 (p = 0.01, 9 polar solvents) and of 0.641 (p = 0.01, 13 non-polar solvents) was considered significant.

**Polar solvent descriptors**

First, the cross-correlations between Dragon and PaDEL PS descriptors was reduced:

``` {r PS correlations removal 1}
PS_corr_DP <-cor(PS_descriptors_training[,8:641], method="pearson")
PS_corr_names_DP <- findCorrelation(PS_corr_DP, cutoff = 0.735, names = TRUE, exact = TRUE)

PS_colnames_DP <- colnames(PS_descriptors_training[8:641])[!(colnames(PS_descriptors_training[8:641]) %in% PS_corr_names_DP)]
```
```{r, echo = FALSE}
cat("Number of descriptors removed:", length(PS_corr_names_DP))
```

Next, the cross-correlations between the reduced number of Dragon and PaDEL descriptors and the remaining PubChem and in-house calculated descriptors were manually removed. 
For this purpose, the cross-correlations between the descriptors were calculated (Pearson two-tailed correlation) and correlation values larger than 0.735 were treated as significant (assigned a "1"), while lower correlations values were treated as insignificant (assigned a "0") .The molecular descriptors self-correlations along the matrix diagonals were also assigned a value of 0 and a binary correlation matrix was obtained.
All descriptors, which were significanly correlated to the in-house and PubChem descriptors were removed.

```{r PS correlations removal 2}
# Binary correlation matrix
PS_colnames_basic_DP <- c(colnames(PS_descriptors_training[2:7]), PS_colnames_DP)

PS_corr_basic_DP <-as.data.frame(cor(PS_descriptors_training[c(PS_colnames_basic_DP)], method="pearson"))
PS_corr_basic_DP[PS_corr_basic_DP <= abs(0.735)] <- 0
PS_corr_basic_DP[PS_corr_basic_DP > abs(0.735)] <- 1
diag(PS_corr_basic_DP) <- 0

# Removal of significantly correlated descriptors
PS_corr_basic_DP <- PS_corr_basic_DP[PS_corr_basic_DP["AmphM_PS"]== 0, PS_corr_basic_DP["AmphM_PS"] == 0]
PS_corr_basic_DP <- PS_corr_basic_DP[PS_corr_basic_DP["Hydrogen.Bond.Acceptor.Count_PS"]== 0, PS_corr_basic_DP["Hydrogen.Bond.Acceptor.Count_PS"] == 0]
```
```{r, echo = FALSE}
cat("Number of descriptors removed:", length(PS_colnames_basic_DP)-length(colnames(PS_corr_basic_DP)),"\n")
cat("Significant correlations left:", sum(PS_corr_basic_DP))
```


**Non-polar solvent descriptors**

An equivalent methodology was employed to remove the significant cross-correlations (Pearson correlation coefficient >= 0.641) withing the NPS molecular descriptors dataset. 

``` {r NPS correlations removal 1}
NPS_corr_DP <-cor(NPS_descriptors_training[,8:684], method="pearson")
NPS_corr_names_DP <- findCorrelation(NPS_corr_DP, cutoff = 0.641, names = TRUE, exact = TRUE)

NPS_colnames_DP <- colnames(NPS_descriptors_training[8:684])[!(colnames(NPS_descriptors_training[8:684]) %in% NPS_corr_names_DP)]
```
```{r, echo = FALSE}
cat("Number of descriptors removed:", length(NPS_corr_names_DP))
```


```{r NPS correlations removal 2}
# Binary correlation matrix
NPS_colnames_basic_DP <- c(colnames(NPS_descriptors_training[2:7]), NPS_colnames_DP)

NPS_corr_basic_DP <-as.data.frame(cor(NPS_descriptors_training[c(NPS_colnames_basic_DP)], method="pearson"))
NPS_corr_basic_DP[NPS_corr_basic_DP <= abs(0.735)] <- 0
NPS_corr_basic_DP[NPS_corr_basic_DP > abs(0.735)] <- 1
diag(NPS_corr_basic_DP) <- 0

# Removal of significantly correlated descriptors
NPS_corr_basic_DP <- NPS_corr_basic_DP[NPS_corr_basic_DP["Molecular.volume..A3._NPS"]== 0, NPS_corr_basic_DP["Molecular.volume..A3._NPS"] == 0]
```
```{r, echo = FALSE}
cat("Number of descriptors removed:", length(NPS_colnames_basic_DP)-length(colnames(NPS_corr_basic_DP)),"\n")
cat("Significant correlations left:", sum(NPS_corr_basic_DP))
```

### Results
A total of **34 polar solvent descriptors** and **24 non-polar solvent descriptors** remained after the above preprocessing steps.

```{r, echo = FALSE}
cat("Polar solvent descriptors")
colnames(PS_corr_basic_DP)
```

```{r, echo = FALSE}
cat("Non-polar solvent descriptors")
colnames(NPS_corr_basic_DP)
```

The results were used to reduce the number of predictors in the descriptor datasets.
```{r}
PS_colnames<-c("PS", colnames(PS_corr_basic_DP))
PS_descriptors<-PS_descriptors[c(PS_colnames)]

NPS_colnames<-c("NPS", colnames(NPS_corr_basic_DP))
NPS_descriptors<-NPS_descriptors[c(NPS_colnames)]
```

```{r, echo = FALSE}
# Export preprocessed datasets 
write.csv(PS_descriptors, "Results/PS descriptors.csv", row.names = FALSE)
write.csv(NPS_descriptors, "Results/NPS descriptors.csv", row.names = FALSE)
```