---
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = TRUE, warning=FALSE, message=FALSE)
```

```{r libraries, echo = FALSE}
library(tidyverse)
library(caret)
library(randomForest)
library(pROC)
```

### Scripts
```{r r scripts}
source("Random forest algorithm.R",local = knitr::knit_global())
source("Predictive performance assessment.R", local = knitr::knit_global())
```

### Dataset
```{r data}
PB_predictors_data <- read.csv("Data/Phase behaviour and descriptors (25 pc step).csv")
```

The dataset was split into a training/testing (`TT_data`) and validation subset comprising 21 and 7 phase diagrams, respectively. Only the training/testing dataset was used to train and evaluate the random forest models employng a reduced set of variables selected based on physico-chemical understanding of the nonaqueous SPC-stabilised systems.
The phase behaviour variable was encoded as a factor.
```{r data splitting}
TT_data <- filter(PB_predictors_data, Set == "Training")
TT_data$Phase.behaviour <- factor(as.character(TT_data$Phase.behaviour))
```

### Random forest models employing a predictor subset selected based on physico-chemical understanding of the nonaqeous SPC-stabilised systems
#### Predictors subset
A list of the 8 compositional and molecular descriptors previously identified as influential for the behaviour of nonaqueous SPC-stabilised systems was defined.
```{r predictors definitions, echo = FALSE}
predictors_pck <- c("PS.concentration", "NPS.concentration", "SPC.concentration", "AmphM_PS","Molecular.volume..A3._PS", "Hydrogen.Bond.Donor.Count_PS","Molecular.volume..A3._NPS", "nCconj_dragon_NPS")
predictors_pck
```
#### Random forest training and evaluation
A random forest model employing the selected predictors was trained and evaluated. 

```{r results dataframe, echo = FALSE}
# Model performance dataframe
model_performance_pck <- data.frame(matrix(ncol = 15, nrow = 20))
colnames(model_performance_pck) <- c("Model", "AUC_training", "Accuracy_training", "Kappa_training", "Sensitivity_training", "Specificity_training", "Precision_training", "Fscore_training", "AUC_testing", "Accuracy_testing", "Kappa_testing", "Sensitivity_testing", "Specificity_testing", "Precision_testing", "Fscore_testing")
```

```{r rf model - GA predictors}
for (i in 1:20) {
  model_performance_pck <- run_random_forest(TT_data, predictors_pck, model_performance_pck, "RF - PCK predictors", i)
}
```

```{r AUC averages, echo=FALSE}
cat("The average AUC achieved by the random forest models (n = 20) on the training data was", round(mean(model_performance_pck$AUC_training),4), "±", round(sd(model_performance_pck$AUC_training),4), "\n")
cat("The average AUC achieved by the random forest models (n = 20) on the testing data was", round(mean(model_performance_pck$AUC_testing), 4), "±", round(sd(model_performance_pck$AUC_testing),4), "\n")
```
```{r export, echo = FALSE}
# Export results
write.csv(model_performance_pck, "Results/Model performance (physico-chemical knowledge).csv", row.names = FALSE)
```